---
title: Introduction to Statistics - Young Researchers Fellowship Program
subtitle: Lecture 1 - Introduction to Statistics & Tabular Data Logic
institute: Laboratorio de Investigación para el Desarrollo del Ecuador
author: Daniel Sánchez Pazmiño
date: 2024-09-01
theme: berlin
date-format: "MMMM YYYY"
format: beamer
incremental: false
fontsize: 10pt
include-in-header:
    - text: |
        \setbeamercolor{background canvas}{bg=white}
        \setbeamertemplate{caption}[numbered]
        \usecolortheme[named=black]{structure}
        \usepackage{tikz}
        \usepackage{pgfplots}
knitr:
    opts_chunk: 
      echo: false
      message: false
      warning: false
---

```{r}
#| label: setup
#| include: false

library(readxl)
library(dplyr)
library(kableExtra)

```

# What is statistics?

- A **methodology** for collecting, analyzing, interpreting, and presenting numerical information.

- A statistic is often referred to as a **numerical fact** or a **piece of data** which describes a particular characteristic of a group of individuals.
    - In the field of statistics, we typically don't refer to individual data points as statistics.

- In several fields, statistics is used as an aid to decision making under uncertainty.

- In a research context, statistics will be needed to understand phenomena, make predictions, and test hypotheses emerging from theory.

- *Statistics is the systematic investigation of the correspondence of theory with the real world*

# Data in statistics

- Because statistics is concerned with information, **data** is often the starting point of any statistical analysis.

- No clear definition of data can possibly satisfy everyone, but we can think of data as a collection of **facts** to be analyzed. 
    - Data is **plural** for **datum**.

- A **dataset** is a collection of data points, which can be organized in a **table**, often about a specific topic, purpose, experiment, study, or context.

# Broad types of data

- Typically, statistics makes a distinction between two broad types of data:

1. **Quantitative** data, which is numerical in nature, meaning it can be measured and expressed in numbers.
    - Discrete data: whole numbers (e.g., number of students in a class).
    - Continuous data: real numbers (e.g., height, weight).

2. **Categorical** or "qualitative" data, which is non-numerical in nature, meaning it cannot be directly measured or expressed in numbers. 
    - Nominal data: categories without order (e.g., colors).
    - Ordinal data: categories with order (e.g., levels of satisfaction).

# How a dataset might look like

```{r}
#| label: example-dataset-supercias

# Load the SUPERCIAS dataset

# supercias_raw <- read_excel("data/directorio_companias_supercias.xlsx", skip = 4)

save(supercias_raw, file = "data/supercias_raw.RData")

```